% Data specification
%
%

This document lists all of the objects in the data specification, and their
fields. The `NULL` column indicates which objects can be empty (*i.e.* no
value is provided). The column `values` indicates the possible values that
this field can take.

Users are encouraged to contribute to the development of this
data specification by suggesting improvements or changes online:
<https://github.com/mangal-wg/mangal-api/issues>.

```{r getApi}
library(rmangal)
options('width' = 150)
url = "http://localhost:8000/"
api = mangalapi(url)
print(api$resources)
```

# Node level

## Taxa

```{r taxa, results='asis'}
kable(whatIs(api, 'taxa'), format="markdown")
```

## Population

```{r pop, results='asis'}
kable(whatIs(api, 'population'), format="markdown")
```

## Item

```{r item, results='asis'}
kable(whatIs(api, 'item'), format="markdown")
```

# Interaction level

## Interaction

```{r interaction, results='asis'}
kable(whatIs(api, 'interaction'), format="markdown")
```

## Network

```{r network, results='asis'}
kable(whatIs(api, 'network'), format="markdown")
```

## Dataset

```{r dataset, results='asis'}
kable(whatIs(api, 'dataset'), format="markdown")
```

# Meta-data

## Trait value

```{r trait, results='asis'}
kable(whatIs(api, 'trait'), format="markdown")
```

## Environmental value

```{r environment, results='asis'}
kable(whatIs(api, 'environment'), format="markdown")
```

## Reference

```{r reference, results='asis'}
kable(whatIs(api, 'reference'), format="markdown")
```

## User

```{r user, results='asis'}
kable(whatIs(api, 'user'), format='markdown')
```
