% `mangal` -- making complex ecological network analysis simpler
% T. Poisot
% Jan. 2014

> This is a *working document* describing `mangal`, a set of `JSON` objects templates to encode ecological networks of virtually any complexity. There are plans to host a pilot database.


```{r, echo=FALSE}
# Prepare the environment
library(devtools)
if(getOption('unzip')=='') options('unzip' = 'unzip')
devtools::install_github(repo='rmangal',username='mangal-wg')
library(rmangal)
mangal_url <- 'http://mangal.uqar.ca/'
api <- mangalapi(mangal_url)
```


# Introduction

{edit}Ecological networks enable ecologists to accommodate the complexity
of natural communities, and to discover mechanisms contributing
to their persistence, stability, resilience, and functioning
[@dunne_network_2006;@bluthgen_what_2008]. Yet, meta-analyses of a
large number of ecological networks are still extremely rare, and most
of the studies comparing several networks {ref} do so within the limit
of particular systems. Networks, as they encode the structure of complex
ecological interactions, have been time and again presented as useful tools
to understand ecosystem properties and dynamics {ref}.  Coming up with a
clear conceptual and mechanistic understanding of the relationships between
the structure of ecological networks and ecosystem properties require to
pool a large quantity of data.

On the other hand, the recent years saw the development of the idea that
network structure is itself a dynamical object, which will change as a function
of environmental conditions and as a result of meta-community processes
{ref}.  Although the *existence* of this variation has been demonstrated,
the reasons for which it happens are much less clearly understood, and
will require a change in focus, from species to populations {ref}. Because
the variability of interactions involve a host of ecological mechanisms,
it is likely that important data mining efforts will be required to fully
understand it. Notably, new approaches based on the replication of networks
over temporal, spatial, and environmental gradients are promising, but require
to have a data structure ready to accomodate the results they will produce.
Beyond just describing the structure of interactions, these data will need to
include informations about environmental context, population characteristics,
and other relevant additional explanatory variables.

In this paper, we (i) establish the need of a data specification serving
as a *lingua franca* among network ecologists, (ii) describe this data
specification. Finally, we (iii) describe `mangal`, a `R` package and
compagnon database, relying on this data specification. We provide some
use cases showing how this new approach makes complex analyzes simpler, and
allows for the integration of new tools to manipulate biodiversity resources.

# Why do we need a data specification?

Ecological networks are (often) stored as their *adjacency matrix* (or as
the quantitative link matrix), that is a series of `0` and `1` indicating,
respectively, the absence and presence of an interaction. This format is
extremely convenient for *use* (as most network analysis packages, *e.g.*
`bipartite`, `betalink`, `foodweb`, require data to be presented this way),
but is extremely inefficient at storing *meta-data*. In most cases, an
adjacency matrix will inform on the identity of species (in cases where
rows and columns headers are present), and the presence or absence of
interactions. If other data about the environment (*e.g.* where the network
wassampled) or the species (*e.g.* the population size, trait distribution,
or other observations) are available, they are most either given in other
files, or as accompanying text. In both cases, making a programmatic link
between interaction data and relevant meta-data is difficult and error-prone.

By contrast, a data specification provides a common language for network
ecologists to interact, and ensure that, regardless of their source, data
can be used in a shared workflow. Most importantly, a data specification
describes how data are *exchanged*. Each group retains the ability to store
the data in the format that is most convenient for in-house use, and only
needs to provide export options (*e.g.* through an API) respecting the
data specification. This approach ensures that *all* data can be used in
meta-analyses, and will in term increase the impact of data {ref}.

# Elements of the data specification

{complete}The data specification is built around the idea that (ecological)
networks are collections of relationships between ecological objects,
each element having particular meta-data associated. In this section,
we detail {complete}. An interactive webpage with the elements of the data
specification can be found online at `http://mangal.uqar.ca./doc/spec/`. The
data specification is implemented as a series of `JSON` schemes,
*i.e.* documents describing how the data should be formatted, and
what each element represent. The schemes can be downloaded from
`https://github.com/mangal-wg/mangal-schemes/releases/tag/1.0`. Rather than giving an exhaustive list of the data specification (which is available online at the aforementionned URL), this section will propose an overview of each element, and of how they interact.

![An overview of the data specification, and the hierarchy between objects. Each box correspond to a level of the data specification. Grey boxes are nodes, blue boxes are interactions and networks, and green boxes are metadata. The **bold** boxes (`dataset`, `network`, `interaction`, `taxa`) are the minimal elements needed to represent a network.](figure-dataspec.pdf)

We propose `JSON` as the most efficient data format for the following
reasons. First, it has emerged as a *de facto* standard for web platform
serving data, and accepting data from users. Second, it allows *validation*
of the data: a `JSON` file can be matched against a scheme, and one can verify
that it is correctly formatted. Finally, `JSON` objects are easily and cheaply
(memory-wise) parsed in the most common programming languages, notably `R`
(equivalent to `list`) and `python` (equivalent to `dict`).

## Node informations

### Taxa

```{r exTaxa, echo=FALSE}
tax = getTaxa(api,'1')
```

Taxa are a taxonomic entity of any level, identified by their name, vernacular
name, and their identifiers in a variety of taxonomic services. Associating
the identifiers of each taxa is important to leverage the power of the new
generation of open data tools, such as `taxize` {ref}. For example, a taxa with an associated *NCBI Taxonomy* identifier can be represented this way:

```json
{
	"name": "`r tax$name`",
	"vernacular": "`r tax$vernacular`",
	"ncbi": "`r as.character(tax$ncbi)`"
}
```

The data specification currently accomodates `ncbi`, `gbif`, `itis` and `bold` identifiers. Correspondances between these and other services can be made through other tools, such as *e.g.* `taxize`.

### Population

### Item

## Network informations

### Interaction

### Network

### Dataset

## Meta-data

### Trait value

### Environmental condition

### User

paternity {ref}

### References

# Use cases

{edit}In this section, we present use cases using the `rmangal` package for
R, to interact with a database implementing this data specification, and
serving data through a `REST`ful API (`http://mangal.uqar.ca/api/v1/`). It
is possible for users to deposit data into this database, through the `R`
package. Data are made available under a *CC-0 Waiver*.

## ...?

## Network beta-diversity

## Connectance and richness relationships

# References
