% `mangal` -- making complex ecological network analysis simpler
% T. Poisot, D. Gravel
% Jan. 2014

> The study of ecological networks is severaly limited by (i) the difficulty
to access data, (ii) the lack of a standardized way to link meta-data with
interactions, and (iii) the disparity of formats in which ecological networks
themselves are represented. To overcome these limitations, we conceived
a data specification for ecological networks. We implemented a database
respecting this standard, and released a R package ( `rmangal`) allowing
users to programmatically access, curate, and deposit data on ecological
interactions. In this article, we show how these tools, in conjunctions with
other frameworks for the programmatic manipulation of open ecological data,
will streamline the analysis process, and improve reproducibility in studies
of ecological networks.

```{r loadStuff, echo=FALSE, message=FALSE, warning=FALSE}
# Prepare the environment
library(devtools)
if(getOption('unzip')=='') options('unzip' = 'unzip')
install_github(repo='rmangal',username='mangal-wg')
library(rmangal)
mangal_url <- 'http://localhost:8000' ## Local version, don't forget to change later!
api <- mangalapi(mangal_url)
```
# Introduction

Ecological networks enable ecologists to accommodate the complexity
of natural communities, and to discover mechanisms contributing to
their persistence, stability, resilience, and functioning. Most of the
"early" studies of ecological networks were focused on understanding
how the structure of interactions within one location affected the
ecological properties of this local community. This led to classical
results, such as the buffering impact of modularity on species loss
@stouffer_compartmentalization_2011, the increase in robustness along with
increases in connectance @dunne_network_2002, and {add_example}. More recently,
new studies introduced the idea that different networks can be meaningfully
compared, either to understand the importance of environmental gradients on
the realisation of ecological interactions [@tylianakis_habitat_2007], or
to understand the mechanisms behind variation in the structure of ecological
networks [@poisot_dissimilarity_2012]. Yet, meta-analyses of a large number
of ecological networks are still extremely rare, and most of the studies
comparing several networks do so within the limit of particular systems
[@schleuning_specialization_2011;@dalsgaard_historical_2013]. In part, this
can be attributed to the limited methods allowing to compare networks in
which no species are in common. However, the severe shortage of data
in the field also restricts the power of large-scale analyses. Indeed, most
of the studies working on several types of interactions focused on comparing
emerging properties [@thebault_does_2008].

An increasing number of approachs are being put forth to *predict* the
structure of ecological networks, either relying on latent variables
[@rohr_modeling_2010] or actual traits [@gravel_inferring_2013]. These
approaches, so as to be adequately calibrated, require easily accessible
data. Comparing the efficiency of different methods will also be facilitated
if there is an homogeneous way of representing ecological interactions,
and the associated metadata. In this paper, we (i) establish the need of a
data specification serving as a *lingua franca* among network ecologists,
(ii) describe this data specification. Finally, we (iii) describe `rmangal`,
a `R` package and compagnon database, relying on this data specification. We
provide some use cases showing how this new approach makes complex analyzes
simpler, and allows for the integration of new tools to manipulate biodiversity
resources.

# Networks need a data specification

Ecological networks are (often) stored as their *adjacency matrix* (or as
the quantitative link matrix), that is a series of `0` and `1` indicating,
respectively, the absence and presence of an interaction. This format is
extremely convenient for *use* (as most network analysis packages, *e.g.*
`bipartite`, `betalink`, `foodweb`, require data to be presented this way),
but is extremely inefficient at *storing* meta-data. In most cases, an
adjacency matrix will inform on the identity of species (in cases where
rows and columns headers are present), and the presence or absence of
interactions. If other data about the environment (*e.g.* where the network
wassampled) or the species (*e.g.* the population size, trait distribution,
or other observations) are available, they are most either given in other
files, or as accompanying text. In both cases, making a programmatic link
between interaction data and relevant meta-data is difficult and error-prone.

By contrast, a data specification provides a common language for network
ecologists to interact, and ensure that, regardless of their source, data
can be used in a shared workflow. Most importantly, a data specification
describes how data are *exchanged*. Each group retains the ability to store
the data in the format that is most convenient for in-house use, and only
needs to provide export options (*e.g.* through an API) respecting the
data specification. This approach ensures that *all* data can be used in
meta-analyses, and will in time increase the impact of data.

# Elements of the data specification

The data specification (Fig. **XX**) is built around the idea that
(ecological) networks are collections of relationships between ecological
objects, each element having particular meta-data associated. In this
section, we detail highlight the way networks are represented in the
`mangal` specification. An interactive webpage with the elements of the data
specification can be found online at `http://mangal.uqar.ca./doc/spec/`. The
data specification is available either at the API root (*e.g.*
`http://mangal.uqar.ca/api/v1/?format=json`), or can be viewed using the
`whatIs` function from the `R` package (see *Supp. Mat. 1*). Rather than
giving an exhaustive list of the data specification (which is available
online at the aforementionned URL), this section will propose an overview
of each element, and of how they interact.

![An overview of the data specification, and the hierarchy between objects. Each box correspond to a level of the data specification. Grey boxes are nodes, blue boxes are interactions and networks, and green boxes are metadata. The **bold** boxes (`dataset`, `network`, `interaction`, `taxa`) are the minimal elements needed to represent a network.](figure-dataspec.pdf)

We propose `JSON` as the most efficient data format for the following
reasons. First, it has emerged as a *de facto* standard for web platform
serving data, and accepting data from users. Second, it allows *validation*
of the data: a `JSON` file can be matched against a scheme, and one can verify
that it is correctly formatted. Finally, `JSON` objects are easily and cheaply
(memory-wise) parsed in the most common programming languages, notably `R`
(equivalent to `list`) and `python` (equivalent to `dict`). For most users,
the format in which data are transmitted will be entirely transparent, as the
interaction will happen within `R` -- as such, knowing how `JSON` objects are
organised is only useful for those who want to interact with  the API directly.

## Node informations

### Taxa

Taxa are a taxonomic entity of any level, identified by their name,
vernacular name, and their identifiers in a variety of taxonomic
services. Associating the identifiers of each taxa is important to leverage
the power of the new generation of open data tools, such as `taxize`
[@chamberlain_taxize:_2013]. The data specification currently accomodates
`ncbi`, `gbif`, `itis`, `eol` and `bold` identifiers.

### Population

A `population` is one observed instance of a `taxa` object. If your
experimental design is replicated through space, then each taxa will have a
`population` object corresponding to each locality. Populations do not have
associated meta-data, but serve as "containers" for `item` objects.

### Item

An `item` is an instance of a population. Items have a `level` argument,
which can be  either `individual` or `population`; this allows to represent
both individual-level networks (*i.e.* there are as many `item`s attached to
a `population` than there were individuals of this `population` sampled),
and population-level networks. When `item` represents a population, it is
possible to give a measure of the size of this population. The notion of
`item` is particularly useful for time-replicated designs: each observation
of a population at a time-point is an `item` with associated `trait` values,
and possibly population size.

## Network informations

### Interaction

An `interaction` links, *a minima*, two `taxa` objects (but can also link pairs
of `population`s or `item`s). The most important attributes of `interaction`s
are the type of interaction (of which we provide a list of possible values,
see *Supp. Mat. 1*), and its `nature`, *i.e.* how it was observed. This
field will help differentiate from direct observations, text mining, and
inference. Note that the `nature` field can also take `absence` as a value;
this will be useful for, *e.g.*, "cafeteria" experiments in which there is
high confidence that the interaction did not happen.

### Network

A `network` is a series of `interaction` object, along with (i) informations
on its spatial position (provided at the latitude and longitude), (ii) the
date of sampling, and (iii) references to measures of environmental conditions. 

### Dataset

A `dataset` is a collection of one or several `network`(s). Datasets also have
a field for `data` and `papers`, both of which are references to bibliographic
or web resources describing, respectively, the source of the data, and the
papers in which these data have been significantly used. Datasets are the
prefered entry point in the resources.

## Meta-data

### Trait value

Objects of type `item` can have associated `trait` values. These consist
in the description of the trait being measured, the value, and the units in
which the measure was taken.

### Environmental condition

Environmental conditions are associated to network. These are defined by
the environmental property measured, its value, and the units.

### References

References are associated to datasets. They accomodate the DOI, JSON or
PubMed identifiers, or a URL. When possible, the DOI should be preferred as
it offers more potential to interact with other on-line tools, such as the
*CrossRef* API.

# Use cases

In this section, we present use cases using the `rmangal` package for `R`,
to interact with a database implementing this data specification, and serving
data through a `REST`ful API (`http://mangal.uqar.ca/api/v1/`). It is possible
for users to deposit data into this database, through the `R` package. Data
are made available under a *CC-0 Waiver*. Detailed informations about how to
upload data are given in the vignettes and manual of the `rmangal` package. So
as to save room in the manuscript, we source each example. The complete `r`
files to reproduce the examples of this section are attached as *Suppl. Mat.*.

The data we use for this example come from @ricciardi_assemblage_2010. These
were previously available on the *InteractionWeb DataBase* as a
single `xls` file. We uploaded them in the `mangal` database at
`http://mangal.uqar.ca/api/v1/dataset/{todo}`.

## Link-species relationships

In the first example, we visualize the relationship between the number of
species and the number of interactions, which @martinez_constant_1992 propose
to be linear (in food webs).

```{r getLS, fig.cap = "Relationship between the number of species and number of interactions in the anemonefish-fish dataset.", dev='pdf', results='asis'}
source('usecases/1_ls.r')
```

Producing this figure requires less than 10 lines of code. The only information
needed is the identifier of the network or dataset, which we suggest should
be reported in publications as: "These data were deposited in the `mangal`
format at `<URL>/api/v1/dataset/<ID>`". This will encourage re-use of the data.

## Network beta-diversity

In the second example, we use the framework of network $\beta$-diversity
[@poisot_dissimilarity_2012] to measure the extent to which networks that are
far apart in space have different interactions. Each network in the dataset
has a latitude and longitude, meaning that it is possible to measure the
geographic distance between two networks.

For each pair of network, we measure the geographic distance (in km.),
the species dissimilarity ($\beta_S$), the network dissimilarity when all
species are present ($\beta_{WN}$), and finally, the network dissimilarity
when only shared species are considered ($\beta_{OS}$).

```{r useBeta, fig.cap="Relationships between the geographic distance between two sites, and the species dissimiliarity, network dissimilarity with all, and only shared, species.", dev='pdf', message=FALSE}
source('usecases/2_beta.r')
```

As shown in *Fig. XX*, while species dissimilarity and overall network
dissimilarity increase when two networks are far apart, this is not the case
for the way common species interact. This suggests that in this system, network
dissimilarity over space is primarily driven by species turnover. The ease
to gather both raw interaction data and associated meta-data make producing
this analysis extremely straigthforward.

## Spatial visualisation of networks

@bascompte_disentangling_2009 proposes an interesting visualisation for
spatialized networks, in which each species is laid out on a map at the
center of mass of its area of occurence; interactions are then drawn between
species, to show how species distribution determines biotic interactions. In
this final use case, we propose to reproduce a similar figure, using the
`RgoogleMaps` package.

```{r useSpace, fig.cap="Spatial plot of a network, using the `maps` and `rmangal` packages. The circle in the inset map show the location of the sites. Each dot in the main map represents a species, with interactions drawn between them.", dev='pdf', message=FALSE}
source('usecases/3_spatial.r')
```
# Conclusions

In this contribution, we presented `mangal`, a data format for the exchange of
ecological networks and associated meta-data. We deployed an online database
with an associated API, relying on this data specification. Finally, we
introduced `rmangal`, a `R` package designed to interact with APIs using the
`mangal` format.

{todo} - <150 wds left for a short conclusion

# References
