% `mangal` -- making ecological network analysis simpler
% Timothée Poisot *et al.* (see complete author list below)
% Feb. 2014

**Author list: **Timothée Poisot (1,2), Benjamin Baiser (3), Jennifer A. Dunne (4,5), Sonia Kéfi (6), François Massol (7,8), Nicolas Mouquet (6), Tamara N. Romanuk (9), Daniel B. Stouffer (1), Spencer A. Wood (10,11), Dominique Gravel (12,2)

1. University of Canterbury, School of Biological Sciences, Christchurch, New Zealand  
2. Québec Centre for Biodiversity Sciences, Montréal (QC), Canada  
3. Department of Wildlife Ecology and Conservation, University of Florida, Gainesville   
4. Sante Fe Institute, 1399 Hyde Park Road, Santa Fe NM 87501   
5. Pacific Ecoinformatics and Computational Ecology Lab, 1604 McGee Ave., Berkeley, CA 94703   
6. Institut des Sciences de l'Évolution, UMR CRNS 5554, Université Montpellier 2, 3405 Montpellier, France  
7. Laboratoire Génétique et Evolution des Populations Végétales, CNRS UMR 8198, Université Lille 1, Bâtiment SN2, F-59655 Villeneuve d’Ascq cedex, France    
8. UMR 5175 CEFE – Centre d'Ecologie Fonctionnelle et Evolutive (CNRS), 1919 Route de Mende, F-34293 Montpellier cedex 05, France    
9. Department of Biology, Dalhousie University   
10. Natural Capital Project, School of Environmental and Forest Sciences, University of Washington, Seattle, WA 98195, USA    
11. Department of Biological Sciences, Idaho State University, Pocatello, ID 83209, USA    
12. Université du Québec à Rimouski, Département de Biologie, 300 Allées des Ursulines, Rimouski (QC) G5L 3A1, Canada  


**Author for correspondence**: `t.poisot@gmail.com`.

\clearpage

**Short title**: Automated retrieval of ecological network data

**Keywords**: R, API, database, open data, ecological networks, species interactions

> The study of ecological networks is severely limited by (i) the difficulty
to access data, (ii) the lack of a standardized way to link meta-data with
interactions, and (iii) the disparity of formats in which ecological networks
themselves are stored and represented. To overcome these limitations, we have designed
a data specification for ecological networks. We implemented a database
respecting this standard, and released a R package (`rmangal`) allowing
users to programmatically access, curate, and deposit data on ecological
interactions. In this article, we show how these tools, in conjunction
with other frameworks for the programmatic manipulation of open ecological
data, streamlines the analysis process and improves replicability and
reproducibility of ecological network studies.

\clearpage

# Introduction

Ecological networks are efficient representations of the complexity of
natural communities, and help discover mechanisms contributing to their
persistence, stability, resilience, and functioning. Most of the "early"
studies of ecological networks were focused on understanding how the structure
of interactions within one location affected the ecological properties
of this local community. They revealed the contribution of 'average'
network properties, such as the buffering impact of modularity on species
loss [@pimm91,@yodz81], the increase in robustness to extinctions along
with increases in connectance [@dunn02], and the fact that organization of
interactions maximizes biodiversity [@bast09]. New studies introduced the idea
that networks can vary from one locality to another. They can be meaningfully
compared, either to understand the importance of environmental gradients
on the presence of ecological interactions [@tyli07], or to understand the
mechanisms behind variation itself [@pois12c;@pois14]. Yet, meta-analyses
of numerous ecological networks are still extremely rare, and most of the
studies comparing several networks do so within the limit of particular systems
[@schl11;@dals13;@pois13;@cham14;@olit14]. The severe shortage of publicly
shared data in the field also restricts the scope of large-scale analyses.

It is possible to predict the structure of ecological networks, either using
latent variables [@rohr10;@eklo13a] or actual trait values [@grav13]. The
calibration of these approaches require accessible data, not only about
the interactions, but about the traits of the species involved. Comparing
the efficiency of different methods is also facilitated if there is a
homogeneous way of representing ecological interactions, and the associated
metadata. In this paper, we (i) establish the need of a data specification
serving as a common language among network ecologists, (ii) describe this
data specification, and (iii) describe `rmangal`, a `R` package and companion
database relying on this data specification. The `rmangal` package allows to
easily deposit and retrieve data about ecological interactions and networks
in a publicly accessible database. We provide use cases showing how this new
approach makes complex analyzes simpler, and allows for the integration of
new tools to manipulate biodiversity resources.

# Networks need a data specification

Ecological networks are (often) stored as an *adjacency matrix* (or as
the quantitative link matrix), that is a series of `0`s and `1`s indicating,
respectively, the absence and presence of an interaction. This format is
extremely convenient for *use* (as most network analysis packages, *e.g.*
`bipartite`, `betalink`, `foodweb`, require data to be presented this way),
but is extremely inefficient at *storing* meta-data. In most cases, an
adjacency matrix provides information about the identity of species (in the cases where rows and
columns headers are present) and the presence or absence of interactions. If
other data about the environment (*e.g.* where the network was sampled)
or the species (*e.g.* the population size, trait distribution, or other
observations) are available, they are often either given in other files
or as accompanying text. In both cases, making a programmatic link between
interaction data and relevant meta-data is difficult and, more importatnly, error-prone.

By contrast, a data specification (*i.e.* a set of precise instructions
detailing how each object should be represented) provides a common language for
network ecologists to interact, and ensures that, regardless of their source,
data can be used in a shared workflow. Most importantly, a data specification
describes how data are *exchanged*. Each group retains the ability to store
the data in the format that is most convenient for in-house use, and only
needs to provide export options (*e.g.* through an API, *i.e.* a programmatic
interface running on a webserver, returning data in response to queries in
a pre-determined language) respecting the data specification. This approach
ensures that *all* data can be used in meta-analyses, and increases the
impact of data [@piwo13]. Data archival also offers additional advantages for
ecology. The aggregation of local observations can reveal large-scale phenomenon
[@reic11], which would be unattainable in the absence of a collaborative
effort. Data archival in databases also prevents data rot and data loss
[@vine14], thus ensuring that data on interaction networks -- which are
typically hard and costly to produce -- continue to be available and usable.

# Elements of the data specification

The data specification introduced here (Fig. 1) is built around the idea that (ecological)
networks are collections of relationships between ecological objects, and each
element has particular meta-data associated with it. In this section, we detail the
way networks are represented in the `mangal` specification. An interactive
webpage with the elements of the data specification can be found online at
`http://mangal.uqar.ca/doc/spec/`. The data specification is available
either at the API root (*e.g.* `http://mangal.uqar.ca/api/v1/?format=json`),
or can be viewed using the `whatIs` function from the `R` package
(see *Supp. Mat. 1*). Rather than giving an exhaustive list of the data
specification (which is available online at the aforementioned URL), this
section serves as an overview of each element, and how they interact.

![An overview of the data specification, and the hierarchy between objects. Every box corresponds to a level of the data specification. Grey boxes are nodes, blue boxes are interactions and networks, and green boxes are metadata. The **bold** boxes (`dataset`, `network`, `interaction`, `taxa`) are the minimal elements needed to represent a network.](figure-dataspec.pdf)

We propose `JSON`, a user-friendly format equivalent to `XML`, as an efficient way to
standardise data representation for two main reasons. First, it has emerged
as a *de facto* standard for web platform serving data, and accepting data
from users. Second, it allows strict *validation* of the data: a `JSON` file can be
matched against a scheme, and one can verify that it is correctly formatted
(this includes the possibility that not all fields are filled, as will
depend on available data). Finally, `JSON` objects are easily and cheaply
(memory-wise) parsed in the most commonly-used programming languages, notably `R`
(equivalent to `list`) and `python` (equivalent to `dict`). For most users,
the format in which data are transmitted is unimportant, as the interaction
happens within `R` -- as such, knowing how `JSON` objects are organized
is only useful for those who want to interact with the API directly. As such, the
`rmangal` package takes care of converting the data into the correct `JSON`
format to upload them in the database.

## Node information

### Taxa

Taxa are a taxonomic entity of any level, identified by their name, vernacular
name, and their identifiers in a variety of taxonomic services. Associating
the identifiers of each taxa allows using the new generation of open data
tools, such as `taxize` [@cham13a], in addition to protecting the database
against taxonomic revisions. The data specification currently has fields
for `ncbi` (National Center for Biotechnology Information), `gbif` (Global
Biodiversity Information Facility), `tsn` (Taxonomic Serial Number, used by
the Integrated Taxonomic Information System), `eol` (Encyclopedia of Life) and
`bold` (Barcode of Life) identifiers. We also provide the taxonomic status,
*i.e.* whether the taxon is a true taxonomic entity, a "trophic species", or a
morphospecies. Taxonomic identifiers can either be added by the contributors,
or will be automatically retrieved during the automated curation routine.

### Item

An `item` is any measured instance of a taxon. Items have a `level` argument,
which can be  either `individual` or `population`; this allows representing
both individual-level networks (*i.e.* there are as many `item`s of a
given taxa as there were individuals of this taxon sampled), and population-level
networks. When `item` represents a population, it is possible to give a measure
of the size of this population. The notion of `item` is particularly useful
for time-replicated designs: each observation of a population at a time-point
is an `item` with associated `trait` values, and possibly population size.

## Network information

All objects described in this sub-section can have a spatial position,
information on the date of sampling, and references to both papers and
datasets.

### Interaction

An `interaction` links two `taxa` objects (but can also link pairs
of `item`s). The most important attributes of `interaction`s are the
type of interaction (of which we provide a list of possible values, see
*Supp. Mat. 1*), and its `ob_type`, *i.e.* how it was observed. This field
helps differentiate direct observations, text mining, and inference. Note
that the `obs_type` field can also take `confirmed absence` as a value;
this is useful for, *e.g.*, "cafeteria" experiments in which there is high
confidence that the interaction did not happen.

### Network

A `network` is a series of `interaction` objects, along with (i) information
on its spatial position (provided at the latitude and longitude), (ii) the
date of sampling, and (iii) references to measures of environmental conditions.

### Dataset

A `dataset` is a collection of one or several `network`(s). Datasets also have
a field for `data` and `papers`, both of which are references to bibliographic
or web resources that describe, respectively, the source of the data and the
papers in which these data have been studied. Datasets or networks
are the preferred entry point into the resources, although in some cases it
can be meaningful to get a list of interactions only.

## Meta-data

### Trait value

Objects of type `item` can have associated `trait` values. These consist
in the description of the trait being measured, the value, and the units in
which the measure was taken. As traits may have been measured at a different
time and/or location that the interaction was, they have fields for time,
latitude and longitude, and references to original publication and original
datasets.

### Environmental condition

Environmental conditions are associated to datasets, networks, and interactions
objects, to allow for both macro and micro environmental conditions. These
are defined by the environmental property measured, its value, and the units.
As traits, they have fields for time, latitude and longitude, and references
to original publication and original datasets.

### References

References are associated to datasets. They accommodate the DOI, JSON or
PubMed identifiers, or a URL. When possible, the DOI is preferred as
it offers more potential to interact with other online tools, such as the
*CrossRef* API.

# Use cases

In this section, we present use cases using the `rmangal` package for `R`,
to interact with a database implementing this data specification, and serving
data through an API (`http://mangal.uqar.ca/api/v1/`). It is possible for users
to deposit data into this database through the `R` package. Note that data are made
available under a *CC-0 Waiver* [@pois13d]. Detailed information
about how to upload data are given in the vignettes and manual of the `rmangal`
package. In addition, the `rmangal` package comes with vignettes explaining
how users can upload their data into the database through `R`.

The data we use for this example come from @ricc10. These data
were previously available on the *InteractionWeb DataBase* as a
single `xls` file. We uploaded them in the `mangal` database at
`http://mangal.uqar.ca/api/v1/dataset/1`. Before running the examples,
users need to install the relevant packages and connect to the database:

```{r loadStuff, echo=TRUE, message=FALSE, warning=FALSE}
# Prepare the environment
library(devtools)
# This line is needed on some linux distributions
if(getOption('unzip')=='') options('unzip' = 'unzip')
# This installs the rmangal package
install_github(repo='rmangal',username='mangal-wg', ref="dev")
install_github(repo='betalink', username='tpoisot', ref="newdoc")
library(betalink)
library(rmangal)
mangal_url <- 'http://localhost:8000/'
api <- mangalapi(mangal_url)
# This packages are required to execute the given examples
library(igraph)
library(plyr)
library(sp)
library(maps)
library(mapdata)
library(RColorBrewer)
# Dataset ID
DSET_ID <- 2
```

## Create taxa and add an interaction

In the first use-case, we will create an interaction between two taxa. We
ask of readers *not* to execute this code as it is, but rather to use it as a
template for their own analyses. A complete, step-by-step guide to upload is
given in the vignettes of the `rmangal` package. Uploading anything require
to have an username and API key, which can be obtained at the following URL:
<http:/mangal.uqar.ca/dashboard/login>. Your API key be generated automatically
after registration. You can use it to connect to the database securely:

~~~ R
api_secure <- mangalapi("http://mangal.io", usr="MyUserName", key="AbcDefIjkL1234")
~~~

The first step is to create two taxa objects, with the species that we
observed interacting:

~~~ R
seal <- list(
   name = "Hydrurga leptonix",
   vernacular = "Leopard seal",
   eol = 328637
)
cod <- list(
   name = "Gadus morhua",
   vernacular = "Atlantic cod"
)
~~~

Now, we will send these two objects in the remote database:

~~~ R
seal <- addTaxa(api_secure, seal)
cod <- addTaxa(api_secure, cod)
~~~

Note that it is suggested to overwrite the local copy of the object, because
the database will *always* send back the remote copy. This makes the syntax
of further addition considerably easier, as we show below. Once the two
objects are created, we can create an interaction between them:

~~~ R
seal_eats_cod  <- list(
   taxa_from = seal,
   taxa_to = cod,
   int_type = "predation",
   obs_type = "observed"
)
~~~

Then using the same approach, we can send this information in the remote
database:

~~~ R
seal_eats_cod <- addInteraction(api_secure, seal_eats_cod)
~~~

To create networks, datasets, etc, one needs follow the same procedure,
as is explained in the vignettes and in the online help.

## Link-species relationships

In the first example, we visualize the relationship between the number of
species and the number of interactions, which @mart92 proposed to be linear
(in food webs).

```{r getLS, fig.cap = "Relationship between the number of species and number of interactions in the anemonefish-fish dataset. Constant connectance refers to the hypothesis that there is a quadratic relationship between these two quantities.", dev='pdf', results='asis'}
# Retrieve the dataset of interest
dataset <- getDataset(api, DSET_ID)

# Get each network in the dataset as a graph object
graphs <- alply(dataset$networks, 1, function(x) toIgraph(api, x))

# Make a data.frame with the number of links and species
ls <- ldply(graphs, function(x) c(S = length(V(x)), L = length(E(x))))
ls$X1 <- aaply(as.numeric(as.vector(ls$X1)), 1,
               function(x) getNetwork(api, x)$name)
colnames(ls)[1] <- 'Network'

# Now plot this dataset
source("suppmat/usecase_ls.r")
```

Getting the data to produce this figure requires less than 10 lines of
code. The only information needed is the identifier of the network or dataset,
which we suggest should be reported in publications as: "These data were
deposited in the `mangal` format at `<URL>/api/v1/dataset/<ID>`" (where
`<URL>` and `<ID>` are replaced by the corresponding values), preferably
in the methods, possibly in the acknowledgements. To encourage data
sharing and its recognition, we encourage users of the database to always cite the
original datasets or publications.

## Network beta-diversity

In the second example, we use the framework of network $\beta$-diversity
[@pois12c] to measure the extent to which networks that are far apart in space
have different interactions. Each network in the dataset has a latitude and
longitude, meaning that it is possible to measure the geographic distance
between two networks.

For each pair of networks, we measure the geographic distance (in km),
the species dissimilarity ($\beta_S$), the network dissimilarity when all
species are present ($\beta_{WN}$), and finally, the network dissimilarity
when only shared species are considered ($\beta_{OS}$).

```{r useBeta, fig.cap="Relationships between the geographic distance between two sites, and the species dissimilarity, network dissimilarity with all species, and network dissimilarity with only shared species.", dev='pdf', message=FALSE}
# We first retrieve all information about the networks
Networks <- alply(dataset$networks, 1, function(x) getNetwork(api, x))

# Extract the lat/lon data
LatLon <- ldply(Networks, function(x) c(name = x$name, lat = x$latitude, lon = x$longitude))
rownames(LatLon) <- LatLon$name
LatLon$lat <- as.numeric(LatLon$lat)
LatLon$lon <- as.numeric(LatLon$lon)
LatLon <- LatLon[,c('lat', 'lon')]

# Then we measure the distances between all pairs of sites
GeoDist <- spDists(as.matrix(LatLon, latlon=TRUE))
colnames(GeoDist) <- rownames(GeoDist) <- rownames(LatLon)
GeoDist <- as.dist(GeoDist)

# Now, we measure the beta-diversity of the networks
names(graphs) <- aaply(names(graphs), 1, function(x) Networks[[x]]$name)
# Finally, we measure the beta-diversity
BetaDiv <- network_betadiversity(graphs)

# We add the geographic distance
BetaDiv$GEO <- GeoDist

# Plotting
source("suppmat/usecase_beta.r")
```

As shown in *Fig. 3*, while species dissimilarity and overall network
dissimilarity increase when two networks are far apart, this is not the case
for the way common species interact. This suggests that in this system, network
dissimilarity over space is primarily driven by species turnover. The ease
to gather both raw interaction data and associated meta-data make conducting
this analysis extremely straightforward.

## Spatial visualization of networks

@basc09a uses an interesting visualization for spatial networks, in which
each species is laid out on a map at the center of mass of its distribution;
interactions are then drawn between species to show how species distribution
determines biotic interactions. In this final use case, we propose to
reproduce a similar figure (*Fig. 4*).

```{r useSpace, fig.cap="Spatial plot of a network, using the `maps` and `rmangal` packages. The circles in the inset map show the location of the sites. Each dot in the main map represents a species, with symbiotic mutualisms drawn between them. The land is in grey.", dev='pdf', message=FALSE}
# We fill a community data matrix
sp_by_site <- llply(graphs, function(x) unlist(V(x)$name))
sp_list <- unique(unlist(sp_by_site))
M <- matrix(0, ncol = length(sp_list), nrow = length(sp_by_site))
colnames(M) <- sp_list
rownames(M) <- names(sp_by_site)
for (site in c(1:length(sp_by_site))) M[names(sp_by_site)[site], sp_by_site[[site]]] = 1

# Next, we get the center position for each species
# (i.e. the mean position of the sites it occurs at)
sp_center <- adply(M, 2, function(x) colMeans(LatLon[names(x)[x > 0], ]))
rownames(sp_center) <- sp_center[, 1]
sp_center <- sp_center[, -1]

# We now create a regional network using betalink::metaweb
Mw <- metaweb(graphs)

# Plot a map
center_point <- colMeans(sp_center)
vcolors <- c(brewer.pal(9, "Set1"), brewer.pal(8, "Set2"))

Layout <- matrix(c(1,2,2,2), 2, 2)
colSize <- c(1.3, 2.0)
rowSize <- c(1.2, 2)
layout(Layout, colSize, rowSize)

# Inset map is number 1
par(mar=c(4.5, 1, 1, 4))
map("worldHires", xlim=c(90,136), ylim=c(-15,15), col="gray90",
      fill=TRUE, resolution=0)
points(center_point[2], center_point[1], pch=1, cex=2, lwd=2)
box()

par(mar=c(4.1, 4.1, 4.1, 4.1))
map("worldHires", xlim=c(124.0,125.1), ylim=c(1.2,1.9), col="gray90",
      fill=TRUE, resolution = 0)
plot(Mw, layout = jitter(as.matrix(LatLon[,c('lon','lat')])),
      rescale = FALSE, add = TRUE, vertex.color = vcolors, vertex.size = 1,
      vertex.label = NA, edge.arrow.size = 0.25, edge.color = 1)
axis(1)
axis(4)
legend("bottomleft", fill = vcolors, legend = V(Mw)$name, inset = 0.02,
      cex = 0.7, bty = "n", ncol=2)
```

# Conclusions

The `mangal` data format will allow researchers to put together dataset
with species interactions and rich meta-data, that are needed to address
emerging questions about the structure of ecological networks. We
deployed an online database with an associated API, relying on this data
specification. Finally, we introduced `rmangal`, a `R` package designed
to interact with APIs using the `mangal` format. We expect that the data
specification will evolve based on the needs and feedback of the community. At
the moment, users are welcome to propose such changes on the project
issue page: <https://github.com/mangal-wg/mangal-schemes/issues>. A
`python` wrapper for the API is also available at
<http://github.com/mangal-wg/pymangal/>. Additionally, there are plans to
integrate this database with *GLOBI*, so that data can be accessed from
multiple sources [@poel14].

**Acknowledgements** This paper was developed during a workshop hosted at the
*Santa Fe Institute*. TP, DBS, and DG acknowledge funding from the Canadian
Institute of Ecology and Evolution. We thank James Caveen for setting up
the database and web server on which the data are hosted. We thank Scott
Chamberlain and one anonymous reviewer for comments on the manuscript.

# References
